---
title: "water access for refugees and americans"
author: "Brendan Sullivan"
date: "October 18, 2019"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

borehole_access <- read_csv("data/BoreholeData_LastAccess_Sep2019.csv")
us_usage <- read_csv("data/domesticwaterusage_bycounty_us.csv")
camp_usage <- read_csv("data/Water_Usage_In_Camps_Most_Recent_Data.csv")

```

## Data exploration

#### Borehole access

1663 observations, high NA population. Variables with around 50% or less NA's could yield good info. 

Numerical variables of interest:

  - Elevation in meters above sea level
  - Depth in m
  - Static water level in meters below ground level
  - Casing diameter in inch
  - Driller's yield in m3/hr
  - Daily pumping time in h

Look at categorical variables like:

  - Country
  - Drilling date
  - Status

```{r explore_borehole_access}

summary(borehole_access)
glimpse(borehole_access)

# restrict dataset to numerical vars of interest and rename
borehole_access_restricted <- borehole_access %>%
  select("Country", 
         "Status",
         "Elevation in meters above sea level", 
         "Depth in m", 
         "Static water level in meters below ground level", 
         "Casing diameter in inch", 
         "Driller's yield in m3/hr", 
         "Daily pumping time in h") %>%
  rename(country = "Country", 
         status = "Status", 
         elevation_m = "Elevation in meters above sea level", 
         depth_m = "Depth in m", 
         static_water_level_m = "Static water level in meters below ground level", 
         casing_diameter_inch = "Casing diameter in inch", 
         drill_yield_m3h = "Driller's yield in m3/hr", 
         daily_pumping_h = "Daily pumping time in h")

# explore dataset
### by country
glimpse(borehole_access_restricted)
(plot_country_pumping <- borehole_access_restricted %>%
  ggplot(aes(country, daily_pumping_h)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))
(plot_country_elev <- borehole_access_restricted %>%
  ggplot(aes(country, elevation_m)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))
(plot_country_depth <- borehole_access_restricted %>%
  ggplot(aes(country, depth_m)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))
(plot_country_staticlevel <- borehole_access_restricted %>%
  ggplot(aes(country, static_water_level_m)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))
(plot_country_casingdiam <- borehole_access_restricted %>%
  ggplot(aes(country, casing_diameter_inch)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))
(plot_country_depth <- borehole_access_restricted %>%
  ggplot(aes(country, depth_m)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))

### explore relationships between borehole variables - depth through daily pumping
(plot_depth_drillyield <- borehole_access_restricted %>%
  ggplot(aes(depth_m, drill_yield_m3h)) +
  geom_point())
(plot_depth_staticwater <- borehole_access_restricted %>%
  ggplot(aes(depth_m, static_water_level_m)) +
  geom_point())
(plot_depth_casing <- borehole_access_restricted %>%
  ggplot(aes(depth_m, casing_diameter_inch)) +
  geom_point())
(plot_depth_pumping <- borehole_access_restricted %>%
  ggplot(aes(depth_m, daily_pumping_h)) +
  geom_point())
(plot_depth_casing <- borehole_access_restricted %>%
  ggplot(aes(depth_m, casing_diameter_inch)) +
  geom_point())
(plot_drill_pump <- borehole_access_restricted %>%
  ggplot(aes(drill_yield_m3h, daily_pumping_h)) +
  geom_point())
(plot_static_drill <- borehole_access_restricted %>%
  ggplot(aes(static_water_level_m, drill_yield_m3h)) +
  geom_point())

```


### Camp usage

796 observations. Same criteria as above - look at variables with approx < 50% NA's. 

Variables of interest:

  - COUNTRY NAME
  - EMERGENCY / POST-EMERGENCY
  - AVERAGE # LITERS OF POTABLE WATER AVAILABLE PER PERSON PER DAY
  - NUMBER OF PERSONS PER USABLE WATER TAP
  -  % OF WATER QUALITY TESTS AT CHLORINATED COLLECTION LOCATIONS WITH FRC IN THE RANGE 0.2-2MG/L AND TURBIDITY <5NTU5
  - NUMBER OF PERSONS PER TOILET/LATRINE
  - NUMBER OF PERSONS PER BATH SHELTER/SHOWER
  - NUMBER OF PERSONS PER HYGIENE PROMOTER
  - REFUGEE POPULATION

```{r explore_camp_usage}

summary(camp_usage)
glimpse(camp_usage)

# restrict to vars of interest and rename
camp_usage_restricted <- camp_usage %>%
  select("COUNTRY NAME", 
         "EMERGENCY / POST-EMERGENCY", 
         "AVERAGE # LITERS OF POTABLE WATER AVAILABLE PER PERSON PER DAY", 
         "NUMBER OF PERSONS PER USABLE WATER TAP", 
         "% OF WATER QUALITY TESTS AT CHLORINATED COLLECTION LOCATIONS WITH FRC IN THE RANGE 0.2-2MG/L AND TURBIDITY <5NTU5", 
         "NUMBER OF PERSONS PER TOILET/LATRINE", 
         "NUMBER OF PERSONS PER BATH SHELTER/SHOWER", 
         "NUMBER OF PERSONS PER HYGIENE PROMOTER", 
         "REFUGEE POPULATION") %>%
  rename(country = "COUNTRY NAME", 
         emergency_status = "EMERGENCY / POST-EMERGENCY", 
         water_per_person_l = "AVERAGE # LITERS OF POTABLE WATER AVAILABLE PER PERSON PER DAY", 
         persons_per_tap = "NUMBER OF PERSONS PER USABLE WATER TAP", 
         quality_tests = "% OF WATER QUALITY TESTS AT CHLORINATED COLLECTION LOCATIONS WITH FRC IN THE RANGE 0.2-2MG/L AND TURBIDITY <5NTU5",
         persons_per_toilet = "NUMBER OF PERSONS PER TOILET/LATRINE", 
         persons_per_shower = "NUMBER OF PERSONS PER BATH SHELTER/SHOWER", 
         persons_per_hygienepromoter = "NUMBER OF PERSONS PER HYGIENE PROMOTER", 
         refugee_population = "REFUGEE POPULATION")

# water quality tests vs other vars
(plot_tests_toilet <- camp_usage_restricted %>%
  ggplot(aes(quality_tests, persons_per_toilet)) +
  geom_point())
(plot_tests_water <- camp_usage_restricted %>%
  ggplot(aes(quality_tests, water_per_person_l)) +
  geom_point())
(plot_tests_tap <- camp_usage_restricted %>%
  ggplot(aes(quality_tests, persons_per_tap)) +
  geom_point())
(plot_tests_shower <- camp_usage_restricted %>%
  ggplot(aes(quality_tests, persons_per_shower)) +
  geom_point())
(plot_tests_hygiene <- camp_usage_restricted %>%
  ggplot(aes(quality_tests, persons_per_hygienepromoter)) +
  geom_point())
(plot_tests_pop <- camp_usage_restricted %>%
  ggplot(aes(quality_tests, refugee_population)) +
  geom_point())

# plotting the distributions of water per person, ppl per tap etc by country would be illustrative when compared with US
# measure water resource crowdedness by country - persons per xyz with additional measure for water per person
(plot_country_tap <- camp_usage_restricted %>%
  ggplot(aes(reorder(country, log(persons_per_tap)), log(persons_per_tap))) +
  geom_jitter(position = position_jitter(w = .2), 
              alpha = .3) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))
(plot_country_water <- camp_usage_restricted %>%
  ggplot(aes(country, water_per_person_l)) +
  geom_jitter(position = position_jitter(w = .2), 
              alpha = .3) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))
(plot_country_pop <- camp_usage_restricted %>%
  ggplot(aes(reorder(country, refugee_population), refugee_population)) +
  geom_jitter(position = position_jitter(w = .2), 
              alpha = .3) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))
(plot_country_toilet <- camp_usage_restricted %>%
  ggplot(aes(reorder(country, persons_per_toilet), persons_per_toilet)) +
  geom_jitter(position = position_jitter(w = .2), 
              alpha = .3) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))
(plot_country_hygiene <- camp_usage_restricted %>%
  ggplot(aes(reorder(country, persons_per_hygienepromoter), persons_per_hygienepromoter)) +
  geom_jitter(position = position_jitter(w = .2), 
              alpha = .3) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))


```


### US usage

with the exception of 4 NA's (out of 3223 total observations) in DO-PSPCp, dataset is complete.

rename variables to make them more intuitive

from the data dictionary this is what they mean 

STATE	State postal abbreviation
STATEFIPS	State FIPS code
COUNTY	County name
COUNTYFIPS	County FIPS code
FIPS	Concatenated State-county FIPS code
YEAR	Year of data=2015
TP-TotPop	Total population of county, in thousands
DO-SSPop	Domestic, self-supplied population, in thousands
DO-WGWFr	Domestic, self-supplied groundwater withdrawals, fresh, in Mgal/d
DO-WSWFr	Domestic, self-supplied surface-water withdrawals, fresh, in Mgal/d
DO-WFrTo	Domestic, total self-supplied withdrawals, fresh, in Mgal/d
DO-SSPCp	Domestic, self-supplied per capita use, in gallons/day [DO-WFrTo/DO-SSPop*1000]
DO-PSDel	Domestic, deliveries from Public Supply, in Mgal/d
DO-PSPCp	Domestic, publicly supplied per capita use, in gallons/day [DO-PSDel/PS-TOPop]
DO-WDelv	Domestic, total use (withdrawals + deliveries), in Mgal/d

Don't need anything with FIPS



```{r explore_us_usage}

summary(us_usage)
glimpse(us_usage)

us_usage_restricted <- us_usage %>%
  select(-c(STATEFIPS, COUNTYFIPS, FIPS)) %>%
  rename(state = STATE, 
         county = COUNTY, 
         year = YEAR, 
         countypop = "TP-TotPop", 
         ss_pop = "DO-SSPop", 
         ss_ground_withdrawals = "DO-WGWFr",
         ss_surface_withdrawals = "DO-WSWFr",
         total_ss_withdrawals = "DO-WFrTo",
         ss_percap_use = "DO-SSPCp",
         del_from_publicsupply = "DO-PSDel",
         pubsupplied_percap_use = "DO-PSPCp",
         totaluse_with_del = "DO-WDelv")

(plot_state_totalsswithdrawals <- us_usage_restricted %>%
  ggplot(aes(reorder(state, total_ss_withdrawals), total_ss_withdrawals)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))

(plot_state_ssgroundwithdrawals <- us_usage_restricted %>%
  ggplot(aes(reorder(state, ss_ground_withdrawals), ss_ground_withdrawals)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))

(plot_state_sssurfacewithdrawals <- us_usage_restricted %>%
  ggplot(aes(reorder(state, ss_surface_withdrawals), ss_surface_withdrawals)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))

(plot_state_delpubsupp <- us_usage_restricted %>%
  ggplot(aes(reorder(state, del_from_publicsupply), del_from_publicsupply)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)))


```